### 2023-09-01 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@5012201741ab7f7cf3b29d541c45d898fad9f315 🚀

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover
risk,\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?},\textbf{Sanchez2022Strategy?}}
these approaches oversimplify the relevant interspecific heterogeneity
in immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{2} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{3} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{4} This creates a disconnect
between coevolutionary theory and most existing ecological frameworks
for mapping spillover risk.

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{5} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{6} or
complexes\textsuperscript{7} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{8} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{9} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{10} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. In their bat reservoirs, coronaviruses evolve through a mix
of host jumps, recombination among disparate lineages, and, to a lesser
degree, co-divergence with their hosts---\textsuperscript{11}a mix of
mechanisms that creates a complex and nonlinear relationship between
host diversity and viral emergence. Working from a recently published
database of bat hosts of betacoronaviruses, we test whether spatial
structure in bat-betacoronavirus coevolution is identifiable at a global
scale. Aiming to explain these patterns, we develop a generalized
framework for applying the GMTC to host-virus interactions, with a
specific emphasis on the potential to create independent coevolutionary
dynamics (and therefore spatial fragmentation in risk) through
heterogeneity. We develop a trivariate risk assessment system that
connects each GMTC mechanism to a quantifiable aspect of host-virus
interactions: (i) viral sharing rates in host communities, representing
the strength of potential interaction between viruses and any one host
(i.e., places where viruses undergo constant host switching may be
coevolutionary coldspots); (ii) the phylogenetic diversity of hosts, as
a proxy for variation in the immunological mechanisms that antagonize
viruses (i.e., the selection mosaic); and (iii) the local uniqueness of
the bat community, representing the potential for viruses to be exposed
to novel host traits (e.g., variation in receptor sequences). Together,
we argue that these can be used to identify and map the evolutionary
drivers that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?}}
Projecting bat and betacoronavirus phylogeny over space
(fig.~\ref{fig:biogeo}), we find support for the idea that bat community
assembly is directly responsible for a global mosaic of viral evolution.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia; (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and Latin America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of multiple
betacoronavirus subgenera, including nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{12}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{11}a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{11,13} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary hosts of the
subgenus \emph{Merbecovirus} (MERS-like viruses);\textsuperscript{12,13}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{14--17} At the coarsest scale, these
heterogeneities are lost, and betacoronavirus biogeography tracks the
deep rifts in bat evolutionary history---but within broad regions, the
component coevolutionary systems may have very different dynamics.

@@

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in their species richness generally track broader patterns of
mammal diversity,\textsuperscript{18} with a striking Neotropical
hotspot (especially in the Amazon basin) and a secondary hotspot
centered in Indochina. These hotspots of bat diversity are generally
presumed to be hotspots of viral adaptive radiation, and therefore areas
of concern for human health.\textsuperscript{11,19} However, the
hotspots of known bat betacoronavirus hosts show a distinct pattern,
with primary hotspots (both in terms of area and higher values) of host
richness situated in southeast Asia, parts of southern Europe, and to a
lesser extent parts of Africa in the -25-0 range of latitudes
(fig.~\ref{fig:richness}; top). Although hundreds of species likely host
undiscovered betacoronaviruses, machine learning predictions have
suggested that these undiscovered reservoirs should follow the same
diversity gradient.\textsuperscript{20} In principle, these hotspots of
locally-diverse, virus-rich bat communities should drive more adaptive
diversification in their viruses.

@@

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South and
Central America have the most evolutionary distinct hosts \emph{and}
viruses, followed by secondary hotspots in southeast Asia and the Rift
Valley region have mostly distinct viruses. Some degree of sampling bias
may contribute to these patterns: for example, the Neotropics are one of
the places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{21--23} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{24}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there is
obvious heterogeneity in betacoronavirus evolutionary distinctiveness
that is distinct from overall bat diversity.

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) include the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to
researchers\textsuperscript{\textbf{Latinne2020Origins?}} (and so have
been sampled most intensively;).\textsuperscript{24} The hotspots of
host richness and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---are almost entirely driven by viral adaptive radiation
through host switching within this clade\textsuperscript{12,20}. In
contrast, the Neotropical hotspot of viral distinctiveness is driven by
isolation by host vicariance. Out of the four main groups of
betacoronaviruses, only merbecoviruses have been found in animals in the
Americas--- an introduction that is generally presumed to be
ancient.\textsuperscript{12,25} While comparatively understudied, New
World merbecoviruses have been found in the ghost-faced bats
(Mormoopidae), Neotropical leaf-nosed bats (Phyllostomidae), and
free-tailed bats (Molossidae).\textsuperscript{14--17} The former two
groups and a clade of the latter are endemic to the Neotropics, while
the explosive adaptive radiations of the phyllostomids are responsible
for the hotspot of bat diversity in the Amazon.\textsuperscript{26}
Together, these clades of New World bats play host to a distinct regime
of betacoronavirus coevolution.

@@

The existence of well-defined cophylogenetic regions suggests that the
bat-betacoronavirus system is spatially fragmented enough to create
divergent coevolutionary trajectories; in turn, this coevolutionary
mosaic may alter the risk of zoonotic emergence. These ideas are,
respectively, supported by the existence of hotspots of viral uniqueness
and the diverse origins of human betacoronaviruses. Together, this
framework points to a predictable relationship between host community
structure and coevolutionary pressure: phylogeographic structure in bat
hosts (and their diverse immune strategies)\textsuperscript{27} creates
a landscape of selective pressure; the trajectory of viruses'
coevolutionary response is, in turn, constrained by their opportunities
for either specialization or diversification through host jumps and
recombination.

@@

Compared to approaches that map emergence risk based only on the number
of known bat hosts of betacoronaviruses, our framework suggests regions
where high viral sharing dominates coevolutionary dynamics---such as
Latin America, or Eurasia above a northing of 30---would pose less of a
relative risk of zoonotic emergence. Nevertheless, areas of high host
uniqueness coupled with high viral sharing (red-to-pink in
fig.~\ref{fig:trivariate}) could create hotspots facilitated by viral
codivergence. Our framework identifies Madagascar, where most bat
species are endemic following evolutionary divergence from sister
species in both African and Asian continents,\textsuperscript{28} as one
such hotspot; interestingly, a recent study\textsuperscript{29} reported
a novel and highly divergent lineage of nobecoviruses from
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus} and
\emph{Rousettus madagascariensis}), again supporting the predictive
power of the coevolutionary framework.

@@

The relationship between the underlying pathogen pool and emergence risk
is mediated by both human-wildlife interfaces (the probability of
spillover) and opportunities for onward horizontal transmission (the
probability that spillovers become epidemics)\textsuperscript{1}. It
must be noted that the assesment of risk based on the GMTC mechanisms
does not account for human presence; for this reason, it represents
``potential'' level of risk, which must be re-evaluated in the light of
human presence. As a proxy for both, we finally overlaid the risk
component from the composite map (see above) with the proportion of
built land, as a proxy for a mix of habitat disturbance, potential for
bat synanthropy or contact with bridge hosts like
livestock,\textsuperscript{30,31} and human population density and
connectivity\textsuperscript{1,32,33} (fig.~\ref{fig:compound}).
Accounting for these factors, most of South America and Europe are at
comparatively lower risk, as--although densely populated--settlements
tend to be in areas with lower potential risk. Conversely, regions like
Malaysia and the northern coast of Australia have a high evolutionary
risk component, but should represent a relatively lower effective risk
due to low human density. However, southeast Asia, the Indian
subcontinent, and scattered hotspots in sub-Saharan Africa are at high
risk due to the overlap between human populations and natural
opportunities for cross-species transmission of betacoronaviruses.

@@

Reassuringly, these predictions correspond to the geographic origins of
the three bat-origin coronaviruses that have recently emerged in human
populations. While available information puts the spillover of
SARS-CoV-2 in a live animal market in Wuhan, China, the ultimate origin
of the virus is almost certainly in a divergent lineage of
sarbecoviruses from Indochina that was poorly characterized prior to the
pandemic.\textsuperscript{21--23} Similarly, the SARS-CoV outbreak began
in Guangdong province in 2002, reaching humans through small carnivore
bridge hosts, but was eventually traced back to a set of likely
progenitor viruses found in cave-dwelling horseshoe bats in Yunnan
province;\textsuperscript{34} nearby, antibody evidence has indicated
human exposure to SARS-like viruses.\textsuperscript{35} MERS-CoV was
first detected in Jordan, but is widespread in camels in East Africa and
the Middle East, and may have reached its bridge host decades earlier
than originally supposed;\textsuperscript{36} as a result, the geography
of the original bat-to-camel transmission is still widely regarded as
uncertain. All of these are broadly consistent with the risk factors we
identify. Notably, India and west Africa are additional hotspots that
have yet to experience the emergence of a bat coronavirus into human
populations, but may still be at risk---particularly given known gaps in
bat surveillance,\textsuperscript{24} and a dense population in both
regions with global connectivity. In any of these regions, surveillance
on viral reservoirs can be paired with targeted monitoring of high-risk
human populations (i.e., those with regular wildlife
contact)\textsuperscript{37} for maximum impact.

@@

Bats emerged around 64 million years ago, and are one of the most
diverse mammalian orders, with more than 1,400 estimated
species.\textsuperscript{\textbf{Peixoto2018SynEco?},\textbf{Simmons2020BatSpe?}}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to humans.\textsuperscript{38} Over
two-thirds of bats are known to be either obligate or facultative
insectivores, therefore actively contributing for agricultural pest
control,\textsuperscript{39,40} and vectors of pathogens that put a risk
on human health;\textsuperscript{41,42} some other species are essential
links in many seed-dispersal networks.\textsuperscript{43} However, many
of these species face a high risk of extinction, particularly given
persecution and killings that sometimes follows from messaging about
their role in disease emergence. Areas where bats, viruses, and humans
co-occur are not always hotspots of risk for human heath; as such,
developing more precise ways to map zoonotic hazards can help bats and
humans coexist safely, and support the conservation of these important
and unique animals.

Here, we propose a simple framework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the once-neglected adaptive radiation of sarbecoviruses
in the Indochinese peninsula. In doing so, it advances ecological theory
beyond the current state of the art for global maps of emergence risk.
For example, previous studies that have used host richness as a proxy
have predicted a high diversity of unsampled bat
viruses,\textsuperscript{19} bat coronaviruses,\textsuperscript{11} and
even specifically betacoronaviruses\textsuperscript{20} in both the
Amazon and southeast Asia. While we find that both regions are
characterized by unique and diverse communities of both hosts and
viruses, our framework is able to identify key differences between the
two systems. We find that the merbecovirus complex in Latin America has
been a unique branch of evolution separate from the rest of the global
pool, but with limited potential for viral diversification--- a finding
that is supported by previous work indicating a higher rate of
codivergence in Latin America.\textsuperscript{11,44} In contrast, in
southeast Asia, host richness and viral distinctiveness are high but
sharing is low; this suggests a different type of evolutionary dynamics
that could generate high local diversity of viruses through host
switching and viral recombination (see \emph{e.g.},\textsuperscript{13}
as well as the discovery of recombinant viruses with genetic material
from both the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{45}Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{20,24} However,
both the evolutionary and ecological aspects of emergence risk are
higher in southeast Asia---a fact that will only become more relevant,
as bats track shifting climates and exchange viruses with other species,
creating a hotspot of elevated cross-species transmission unique to the
region.\textsuperscript{32,46}

Our trivariate additive mapping of components of risk
(fig.~\ref{fig:trivariate}) aims to elicit the complexity of spatial
cross-species transmission risk beyond the mere presence or absence of
the pathogen host in a specific location. By considering coevolutionary
factors such as viral sharing and host uniqueness, we suggest insights
that can aid in identifying potential locations for surveillance of
betacoronavirus circulation and assessing the risk of cross-species
transmission to other mammals. In communities characterized by diverse
but unique host populations, with limited viral sharing between them, we
could encounter viruses that specialize in targeting the immune system
of specific hosts. This implies a low likelihood of infecting novel
hosts but, once locally introduced into a new host (either a new
species, or an immunologically naïve population), the specialized virus
could spread relatively easily due to encountering little immune
resistance.\textsuperscript{\textbf{Plowright2011Urban?}} With the right
combination of viral traits, such as low disease-induced mortality or
high transmission rate, this could lead to successfully spread within
the new host community. However, while high adaptation to a specific
host can be advantageous, it may also lead to maladaptation when the
pathogen encounters a new unsuitable host, potentially resulting in its
extinction.

>>>>

Disease emergence is complex, and is driven not only by animal-human
contact, but also by the underlying evolutionary dynamics in viral
reservoirs.\textsuperscript{1} Although host richness is often used as a
superficial proxy for spillover
risk,\textsuperscript{2,3,\textbf{Ruiz-Aravena2022Ecological?}} these
approaches oversimplify the relevant interspecific heterogeneity in
immunology, behavior, and other traits, and therefore overlook unique
host pools that allow for the rapid evolution of highly divergent
viruses.\textsuperscript{4} In the case of generalist pathogens like
betacoronaviruses, there is conceptual and empirical support to the idea
that these community-level mechanisms are even more
important,\textsuperscript{5} particularly given that cross-species
transmission may, as a rule, structure viral evolution more than
co-divergence with hosts.\textsuperscript{6} This creates a disconnect
between coevolutionary theory and most existing ecological frameworks
for mapping spillover risk.

The geographic mosaic theory of coevolution (GMTC) attempts to
explicitly connect microevolutionary dynamics to the macroecology and
biogeography of symbiotic interactions.\textsuperscript{7} The GMTC
posits that coevolutionary processes among pairs\textsuperscript{8} or
complexes\textsuperscript{9} of species are structured in space by the
rippling effects of abiotic conditions onto evolutionary mechanisms,
giving rise to fragmented systems with different ecologies over large
spatial extents.\textsuperscript{10} The GMTC predicts a spatial
fragmentation of coevolutionary dynamics under the joint action of three
processes:\textsuperscript{11} coevolutionary hot- and coldspots, which
appear when the intensity of \emph{interaction} (in terms of reciprocal
fitness consequences) varies spatially; selection mosaics, wherein the
intensity of \emph{selection} varies across space, driven by both the
biotic complexity of the community (locally diverse hosts and viruses
are more biotically complex) and the local favorability of the
environment;\textsuperscript{12} and trait remixing, which occurs when
coevolutionary dynamics change when community-level \emph{functional
traits} change through meta-community dynamics.

Here, we apply the GMTC to explore and explain the global biogeography
of betacoronaviruses, the group that includes SARS-CoV, MERS-CoV, and
SARS-CoV-2. In their bat reservoirs, coronaviruses evolve through a mix
of host jumps, recombination among disparate lineages, and, to a lesser
degree, co-divergence with their hosts---\textsuperscript{2}a mix of
mechanisms that creates a complex and nonlinear relationship between
host diversity and viral emergence. Working from a recently published
database of bat hosts of betacoronaviruses, we test whether spatial
structure in bat-betacoronavirus coevolution is identifiable at a global
scale. Aiming to explain these patterns, we develop a generalized
framework for applying the GMTC to host-virus interactions, with a
specific emphasis on the potential to create independent coevolutionary
dynamics (and therefore spatial fragmentation in risk) through
heterogeneity. We develop a trivariate risk assessment system that
connects each GMTC mechanism to a quantifiable aspect of host-virus
interactions: (i) viral sharing rates in host communities, representing
the strength of potential interaction between viruses and any one host
(i.e., places where viruses undergo constant host switching may be
coevolutionary coldspots); (ii) the phylogenetic diversity of hosts, as
a proxy for variation in the immunological mechanisms that antagonize
viruses (i.e., the selection mosaic); and (iii) the local uniqueness of
the bat community, representing the potential for viruses to be exposed
to novel host traits (e.g., variation in receptor sequences). Together,
we argue that these can be used to identify and map the evolutionary
drivers that---in conjunction with transmission processes (e.g., viral
prevalence in reservoirs and animal-human contact rates)--- determine
disease emergence risk.

@@

Most previous work has assumed that the presence or richness of key
groups of bat hosts are predictive of coronavirus
diversity.\textsuperscript{\textbf{Anthony2017GloPat?},\textbf{Ruiz-Aravena2022EcoEvo?}}
Projecting bat and betacoronavirus phylogeny over space
(fig.~\ref{fig:biogeo}), we find support for the idea that bat community
assembly is directly responsible for a global mosaic of viral evolution.
The distinct groupings (represented by different colors, symbolizing
positions in a subspace formed by the first two phylogenetic principal
components) are essentially equivalent between the two groups, and can
be coarsely delineated as (1) south and southeast Asia; (2) east Asia
(including northern China), west Asia, and the Mediterranean coast; (3)
Eurasia above a northing of 40; and (4) Africa and Latin America. In
some cases, this diverges from expectations about coronavirus
biogeography: for example, previous work has rarely flagged India as a
region of interest, but for both bats and betacoronaviruses, the
subcontinent falls into the same regions as the southeast Asian
peninsula (and indeed, the region is home to known bat hosts of multiple
betacoronavirus subgenera, including nobecoviruses, sarbecoviruses, and
merbecoviruses).\textsuperscript{13}

@@

Overall, these results suggest that the boundaries of bat and
betacoronavirus biogeographic regions are broadly consistent at a global
scale; perfect matching between the biogeographic regions would have
indicated that the signal of virus distribution is fully predicted by
bat hosts ranges. Areas for which the biogeographic regions for bats and
betacoronaviruses differ are primarily (i) southeast Asia and southern
China, and (ii) the Arabian peninsula, which are both regions where
zoonotic transmission has been documented (potentially driving a unique
level of viral sampling effort that generates these patterns). These
spatially limited mismatches nonwithstanding, the large level of
congruence may be surprising, given that cross-species transmission may
play a stronger role in coronavirus diversification than
cospeciation---\textsuperscript{2}a property that would theoretically
allow for substantial broad divergence in their biogeography. However,
host jumps at the family level or higher are relatively rare and
significant events in coronavirus evolutionary
history;\textsuperscript{2,14} as a result, the mosaic of
betacoronavirus phylogeography is assembled from a set of overlapping
smaller coevolutionary systems, superimposed in space and filtered by
the importance of different subgroups in local host communities. For
example, the most speciose and cosmopolitan family of bats, the vesper
bats (Vespertilionidae), are considered the primary hosts of the
subgenus \emph{Merbecovirus} (MERS-like viruses);\textsuperscript{13,14}
but in the Americas, where merbecoviruses are the only lineage present,
they have only been found in other bat taxa (\emph{e.g.}, Molossidae,
Phyllostomidae).\textsuperscript{15--18} At the coarsest scale, these
heterogeneities are lost, and betacoronavirus biogeography tracks the
deep rifts in bat evolutionary history---but within broad regions, the
component coevolutionary systems may have very different dynamics.

@@

Bats, the second most diverse groups of mammals, are found worldwide;
gradients in their species richness generally track broader patterns of
mammal diversity,\textsuperscript{19} with a striking Neotropical
hotspot (especially in the Amazon basin) and a secondary hotspot
centered in Indochina. These hotspots of bat diversity are generally
presumed to be hotspots of viral adaptive radiation, and therefore areas
of concern for human health.\textsuperscript{2,20} However, the hotspots
of known bat betacoronavirus hosts show a distinct pattern, with primary
hotspots (both in terms of area and higher values) of host richness
situated in southeast Asia, parts of southern Europe, and to a lesser
extent parts of Africa in the -25-0 range of latitudes
(fig.~\ref{fig:richness}; top). Although hundreds of species likely host
undiscovered betacoronaviruses, machine learning predictions have
suggested that these undiscovered reservoirs should follow the same
diversity gradient.\textsuperscript{21} In principle, these hotspots of
locally-diverse, virus-rich bat communities should drive more adaptive
diversification in their viruses.

@@

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South and
Central America have the most evolutionary distinct hosts \emph{and}
viruses, followed by secondary hotspots in southeast Asia and the Rift
Valley region have mostly distinct viruses. Some degree of sampling bias
may contribute to these patterns: for example, the Neotropics are one of
the places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{22--24} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{25}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there is
obvious heterogeneity in betacoronavirus evolutionary distinctiveness
that is distinct from overall bat diversity.

Overall, these patterns recapitulate the evolutionary history of both
the order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe
bats (Rhinolophidae) include the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}), the group of pandemic threats that have
been of the greatest interest to
researchers\textsuperscript{\textbf{Latinne2020Origins?}} (and so have
been sampled most intensively;).\textsuperscript{25} The hotspots of
host richness and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---are almost entirely driven by viral adaptive radiation
through host switching within this clade\textsuperscript{13,21}. In
contrast, the Neotropical hotspot of viral distinctiveness is driven by
isolation by host vicariance. Out of the four main groups of
betacoronaviruses, only merbecoviruses have been found in animals in the
Americas--- an introduction that is generally presumed to be
ancient.\textsuperscript{13,26} While comparatively understudied, New
World merbecoviruses have been found in the ghost-faced bats
(Mormoopidae), Neotropical leaf-nosed bats (Phyllostomidae), and
free-tailed bats (Molossidae).\textsuperscript{15--18} The former two
groups and a clade of the latter are endemic to the Neotropics, while
the explosive adaptive radiations of the phyllostomids are responsible
for the hotspot of bat diversity in the Amazon.\textsuperscript{27}
Together, these clades of New World bats play host to a distinct regime
of betacoronavirus coevolution.

@@

The existence of well-defined cophylogenetic regions suggests that the
bat-betacoronavirus system is spatially fragmented enough to create
divergent coevolutionary trajectories; in turn, this coevolutionary
mosaic may alter the risk of zoonotic emergence. These ideas are,
respectively, supported by the existence of hotspots of viral uniqueness
and the diverse origins of human betacoronaviruses. Together, this
framework points to a predictable relationship between host community
structure and coevolutionary pressure: phylogeographic structure in bat
hosts (and their diverse immune strategies)\textsuperscript{28} creates
a landscape of selective pressure; the trajectory of viruses'
coevolutionary response is, in turn, constrained by their opportunities
for either specialization or diversification through host jumps and
recombination.

@@

Compared to approaches that map emergence risk based only on the number
of known bat hosts of betacoronaviruses, our framework suggests regions
where high viral sharing dominates coevolutionary dynamics---such as
Latin America, or Eurasia above a northing of 30---would pose less of a
relative risk of zoonotic emergence. Nevertheless, areas of high host
uniqueness coupled with high viral sharing (red-to-pink in
fig.~\ref{fig:trivariate}) could create hotspots facilitated by viral
codivergence. Our framework identifies Madagascar, where most bat
species are endemic following evolutionary divergence from sister
species in both African and Asian continents,\textsuperscript{29} as one
such hotspot; interestingly, a recent study\textsuperscript{30} reported
a novel and highly divergent lineage of nobecoviruses from
Madagascar-endemic pteropid bat species (\emph{Pteropus rufus} and
\emph{Rousettus madagascariensis}), again supporting the predictive
power of the coevolutionary framework.

@@

The relationship between the underlying pathogen pool and emergence risk
is mediated by both human-wildlife interfaces (the probability of
spillover) and opportunities for onward horizontal transmission (the
probability that spillovers become epidemics)\textsuperscript{1}. It
must be noted that the assesment of risk based on the GMTC mechanisms
does not account for human presence; for this reason, it represents
``potential'' level of risk, which must be re-evaluated in the light of
human presence. As a proxy for both, we finally overlaid the risk
component from the composite map (see above) with the proportion of
built land, as a proxy for a mix of habitat disturbance, potential for
bat synanthropy or contact with bridge hosts like
livestock,\textsuperscript{31,32} and human population density and
connectivity\textsuperscript{1,33,34} (fig.~\ref{fig:compound}).
Accounting for these factors, most of South America and Europe are at
comparatively lower risk, as--although densely populated--settlements
tend to be in areas with lower potential risk. Conversely, regions like
Malaysia and the northern coast of Australia have a high evolutionary
risk component, but should represent a relatively lower effective risk
due to low human density. However, southeast Asia, the Indian
subcontinent, and scattered hotspots in sub-Saharan Africa are at high
risk due to the overlap between human populations and natural
opportunities for cross-species transmission of betacoronaviruses.

@@

Reassuringly, these predictions correspond to the geographic origins of
the three bat-origin coronaviruses that have recently emerged in human
populations. While available information puts the spillover of
SARS-CoV-2 in a live animal market in Wuhan, China, the ultimate origin
of the virus is almost certainly in a divergent lineage of
sarbecoviruses from Indochina that was poorly characterized prior to the
pandemic.\textsuperscript{22--24} Similarly, the SARS-CoV outbreak began
in Guangdong province in 2002, reaching humans through small carnivore
bridge hosts, but was eventually traced back to a set of likely
progenitor viruses found in cave-dwelling horseshoe bats in Yunnan
province;\textsuperscript{35} nearby, antibody evidence has indicated
human exposure to SARS-like viruses.\textsuperscript{36} MERS-CoV was
first detected in Jordan, but is widespread in camels in East Africa and
the Middle East, and may have reached its bridge host decades earlier
than originally supposed;\textsuperscript{37} as a result, the geography
of the original bat-to-camel transmission is still widely regarded as
uncertain. All of these are broadly consistent with the risk factors we
identify. Notably, India and west Africa are additional hotspots that
have yet to experience the emergence of a bat coronavirus into human
populations, but may still be at risk---particularly given known gaps in
bat surveillance,\textsuperscript{25} and a dense population in both
regions with global connectivity. In any of these regions, surveillance
on viral reservoirs can be paired with targeted monitoring of high-risk
human populations (i.e., those with regular wildlife
contact)\textsuperscript{38} for maximum impact.

@@

Bats emerged around 64 million years ago, and are one of the most
diverse mammalian orders, with more than 1,400 estimated
species.\textsuperscript{\textbf{Peixoto2018SynEco?},\textbf{Simmons2020BatSpe?}}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to humans.\textsuperscript{39} Over
two-thirds of bats are known to be either obligate or facultative
insectivores, therefore actively contributing for agricultural pest
control,\textsuperscript{40,41} and vectors of pathogens that put a risk
on human health;\textsuperscript{42,43} some other species are essential
links in many seed-dispersal networks.\textsuperscript{44} However, many
of these species face a high risk of extinction, particularly given
persecution and killings that sometimes follows from messaging about
their role in disease emergence. Areas where bats, viruses, and humans
co-occur are not always hotspots of risk for human heath; as such,
developing more precise ways to map zoonotic hazards can help bats and
humans coexist safely, and support the conservation of these important
and unique animals.

Here, we propose a simple framework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the once-neglected adaptive radiation of sarbecoviruses
in the Indochinese peninsula. In doing so, it advances ecological theory
beyond the current state of the art for global maps of emergence risk.
For example, previous studies that have used host richness as a proxy
have predicted a high diversity of unsampled bat
viruses,\textsuperscript{20} bat coronaviruses,\textsuperscript{2} and
even specifically betacoronaviruses\textsuperscript{21} in both the
Amazon and southeast Asia. While we find that both regions are
characterized by unique and diverse communities of both hosts and
viruses, our framework is able to identify key differences between the
two systems. We find that the merbecovirus complex in Latin America has
been a unique branch of evolution separate from the rest of the global
pool, but with limited potential for viral diversification--- a finding
that is supported by previous work indicating a higher rate of
codivergence in Latin America.\textsuperscript{2,45} In contrast, in
southeast Asia, host richness and viral distinctiveness are high but
sharing is low; this suggests a different type of evolutionary dynamics
that could generate high local diversity of viruses through host
switching and viral recombination (see \emph{e.g.},\textsuperscript{14}
as well as the discovery of recombinant viruses with genetic material
from both the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{46}Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{21,25} However,
both the evolutionary and ecological aspects of emergence risk are
higher in southeast Asia---a fact that will only become more relevant,
as bats track shifting climates and exchange viruses with other species,
creating a hotspot of elevated cross-species transmission unique to the
region.\textsuperscript{33,47}

Our trivariate additive mapping of components of risk
(fig.~\ref{fig:trivariate}) aims to elicit the complexity of spatial
cross-species transmission risk beyond the mere presence or absence of
the pathogen host in a specific location. By considering coevolutionary
factors such as viral sharing and host uniqueness, we suggest insights
that can aid in identifying potential locations for surveillance of
betacoronavirus circulation and assessing the risk of cross-species
transmission to other mammals. In communities characterized by diverse
but unique host populations, with limited viral sharing between them, we
could encounter viruses that specialize in targeting the immune system
of specific hosts. This implies a low likelihood of infecting novel
hosts but, once locally introduced into a new host (either a new
species, or an immunologically naïve population), the specialized virus
could spread relatively easily due to encountering little immune
resistance.\textsuperscript{48} With the right combination of viral
traits, such as low disease-induced mortality or high transmission rate,
this could lead to successfully spread within the new host community.
However, while high adaptation to a specific host can be advantageous,
it may also lead to maladaptation when the pathogen encounters a new
unsuitable host, potentially resulting in its extinction.
